---
title: "Data Manipulation"
output: github.document
---

```{r setup, echo=FALSE}
library(tidyverse)
```

```{r, load data}

litters_data = read_csv("./data_import_examples/FAS_litters.csv")
litters_data = janitor::clean_names(litters_data)

```

##Selecting a subset of columns
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Helper functions with select()

```{r}
select(litters_data, starts_with("gd"))
select(litters_data, ends_with("weight"))
select(litters_data, litter_number, pups_survive, everything())
```

## `filter`ing data

* `gd_of_birth == 20`
* `pups_born_alive >= 2`
* `pups_survive != 4`
* `!(pups_survive == 4)`
* `group %in% c("Con7", "Con8")`
* `group == "Con7" & gd_of_birth == 20`

Let's try one or two... 
```{r}
filter(litters_data, gd_of_birth == 20)
```

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

A very common filtering step requires you to omit missing observations. You *can* do this with `filter`, but I recommend using `drop_na` from the `tidyr` package:

* `drop_na(litters_data)` will remove any row with a missing value
* `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing. 

Filtering can be helpful for limiting a dataset to only those observations needed for an analysis. However, I recommend against the creation of many data subsets (e.g. one for each group). This can clutter up your workspace, and we'll see good tools for the analysis of subsets before long.

### `mutate`

Sometimes you need to select columns; sometimes you need to change them or create new ones. You can do this using `mutate`. 

The example below creates a new variable measuring the difference between `gd18_weight` and `gd0_weight` and modifies the existing `group` variable.

```{r}
#litter_data2 =
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group),
 # wt_gain_kg = wt_gain * 2.2
)
```
A few things in this example are worth noting:

* Your new variables can be functions of old variables
* New variables appear at the end of the dataset in the order that they are created
* You can overwrite old variables
* You can create a new variable and immediately refer to (or change) it

Creating a new variable that does exactly what you need can be a challenge; the more functions you know about, the easier this gets. 

### `arrange`

In comparison to the preceding, arranging is pretty straightforward. You can arrange the rows in your data according to the values in one or more columns:

```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

You can also sort in descending order if you'd like.
```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)
```
### `%>%`

We've seen several commands you will use regularly for data manipulation and cleaning. You will rarely use them in isolation. For example, suppose you want to load the data, clean the column names, remove `pups_survive`, and create `wt_gain`. There are a couple of options for this kind of multi-step data manipulation:

* define intermediate datasets (or overwrite data at each stage)
* nest function calls

The following is an example of the first option:

```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

Below, we try the second option:

```{r}
litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )
litters_data_clean
```

These are both confusing and bad: the first gets confusing and clutters our workspace, and the second has to be read inside out.

Piping solves this problem. It allows you to turn the nested approach into a sequential chain by passing the result of one function call as an argument to the next function call:

```{r}
litters_data = 
  read_csv("./data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>% #not specifying dataset here; that is the key to piping. Piping creates a new tibble automatically
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)
litters_data
```
Use Ctrl+Shift+M for ` %>% `